Learning to Synthesize Rules for Network Intrusion Detectors

### 2020-01-04 paper/paper.tex
another pass

\Mar{estou aqui-----------------}

@@

Figure~\ref{fig:synflood-example} shows an example rule of
Suricata~\cite{suricata}, a popular open-source \nids{} maintained by
the Open Information Security Foundation (OISF)~\cite{oisf}. A
Suricata rule is divided in three parts---action, header, and rule
options~\cite{suri-rule-format}. The action part appears as the first
word in the rule description. An action denotes the task that needs to
be executed if the rule pattern is satisfied. In this example, a
message will be sent to system administrators if the rule pattern is
satisfied. The header comes after the action in the rule
description. It restricts the information flow covered by the
rule. For this rule, the header is \CodeIn{tcp \$HOME\_NET any ->
  \$EXTERNAL\_NET any}. It instructs Suricata to inspect \CodeIn{tcp}
traffic flowing from any port in the home network to any other address
outside the home network. The variables \CodeIn{\$HOME\_NET} and
\CodeIn{\$EXTERNAL\_NET} are configurable. The rule options come after
the header. It is a semi-colon-separated sequence of key-value
pairs. Options serve to 1) document the analyzed traffic (\eg{},
``msg'', ``classtype'', and ``sid'') and to 2) describe
characteristics of the attack (\eg, ``flags'', ``threshold'', and
``content''). \tname{} produces options of the second kind, which are
\emph{relevant} to capturing the attack.

@@

Suricata supports a total of \totoptions\ distinct options covering
\numproto\ distinct protocols. A total of 9 of these options are for
documentation.\MyComment{For instance, the purpose of rule option \CodeIn{msg} is to print on
the output an informative message indicating what kind of intrusion
has been observed.} Figure~\ref{fig:distribution-rules-protocol} shows
the distributions of these options per protocol. The sum of
the number of protocol options is higher than \totoptions\ as some
options is shared across different protocols.

Table~\ref{table:rules} shows a sample of the options of
Suricata. Column ``\#'' shows the option number, column ``Name'' shows
the name of the option, and column
``Description'' presents a short description of the option purpose. We
found that the option \CodeIn{content} is particularly prevalent in
rules from public databases (see
Figure~\ref{fig:distribution-contents}). This option takes a sequence
of bytes as argument (\eg{}, \CodeIn{content:''SELECT ''}) and checks
if that sequence is present in the payload of the message. Detailed
description of the Suricata rules can be found
elsewhere~\cite{suri-rule-format}.

@@

\begin{figure*}[h!]
\centering
\includegraphics[scale=0.3]{figs/ICMP-packet-structure.png}
\caption{Layout of an ICMP packet.}
\label{fig:icmp-packet-layout}
\end{figure*}

There exists a correspondence between network packets and most rule
options. Let's consider the ICMP protocol as an
example. Figure~\ref{fig:icmp-packet-layout} shows the layout of an
ICMP packet. The packet fields ``Flags'' and ``Time to Live'',
declared in the IP header, are mapped to the options \CodeIn{flagbits}
and \CodeIn{ttl}, respectively. Likewise, the fields ``Type'' and
``Code'', declared in the ICMP header, are mapped to the options
\CodeIn{itype} and \CodeIn{icode}. (Table~\ref{table:rules} lists some
example rule options.) The values of the options ``ttl'', ``itype'',
and ``icode'' are the same as the value of the corresponding fields in
the ICMP packet. For the field ``Flags'' the value of the
corresponding option is encoded in its bitvector. For example, if the
``Flags'' field stores the value 1010, the corresponding rule option
will be \CodeIn{flagbits:RM}, as each position of the vector encodes a
different character in the option \CodeIn{flagbits}.

The value of most options can be obtained through transformation
functions, as those described above. However, this is not the case for
the option \CodeIn{threshold}, which is typically used in
denial-of-service attacks. Conceptually, that option operates over
sequence of messages as opposed to a single message. More precisely,
to trigger an action on a rule containing that option the \nids\ tool
needs to monitor several messages.

@@

@@

%% Explain how \tname\ behaves on this attack

In the following, we illustrate how \tname{} proceeds for obtaining
the rule for the SYN Flood attack. The main challenge in synthesizing
rules is to discover the set of options that enables the \nids\ to
capture the malicious traffic, but ignore the benign traffic.  The
other parts of the rule (\ie{}, the action and the header) are either
inferred or left constant. \tname{} sets the action header to
\CodeIn{alert} and sets the header to \CodeIn{<proto> any any -> any
  any}, indicating that it infers the target protocol and uses the
most general flow description possible, \ie{}, it instructs the
\nids\ to analyze the traffic flowing from/to any address and port.
We understand that this is domain-specific information and conjecture
that the system administrator would know how to modify the constant
parts.

\tname{} proceeds as follows to synthesize rules for the SYN Flood
attack. First, it analyzes the positive traffic and extracts options
expressed in that traffic. For this example, it finds the following
set of options:

@@

Note that a rule that uses this set of options would capture the
positive traffic as these options originated from the positive
traffic. Unfortunately, such rule would be overspecified. It would
unlikely capture different manifestations of the same
attack. Informally, the rule includes ``too many'' characteristics of
the malicious traffic used to define it. To address this overfitting
issue, \tname{} minimizes the rule preserving the invariant that
plausible rules should only capture the positive traffic. \tname{}
potentially finds several minimal rules. The list below shows the
options for the five shortest rules it produces for this attack.

@@

>>>>

@@

Figure~\ref{fig:synflood-example} shows an example rule of
Suricata~\cite{suricata}, a popular open-source \nids{} maintained by
the Open Information Security Foundation (OISF)~\cite{oisf}. A
Suricata rule is divided in three parts---action, header, and rule
options~\cite{suri-rule-format}. The action part appears as the first
word in the rule description. An action denotes the task that needs to
be executed if the rule pattern is satisfied. In this example, a
message will be sent to system administrators if the rule pattern is
satisfied. The header comes after the action in the rule
description. It restricts the information flow covered by the
rule. For this rule, the header is \CodeIn{tcp \$HOME\_NET any ->
  \$EXTERNAL\_NET any}. It instructs Suricata to inspect \CodeIn{tcp}
traffic flowing from any port in the home network to any other address
outside the home network. The variables \CodeIn{\$HOME\_NET} and
\CodeIn{\$EXTERNAL\_NET} are configurable. The rule options come after
the header. It is a semi-colon-separated sequence of key-value
pairs. Rule options serve to document the analyzed traffic (\eg{},
``msg'', ``classtype'', and ``sid'') and to characterize the attack
pattern (\eg, ``flags'', ``threshold'', and ``content''). \tname{}
produces options of the second kind, which are \emph{relevant} to
capturing the attack.

Table~\ref{table:rules} shows a sample of the options of
Suricata. Column ``\#'' shows the option number, column ``Name'' shows
the name of the option, and column ``Description'' presents a short
description of the option purpose. We found that the option
\CodeIn{content} is particularly prevalent in rules from public
databases\MyComment{ (see Figure~\ref{fig:distribution-contents})}. This
option takes a sequence of bytes as argument (\eg{},
\CodeIn{content:''SELECT ''}) and checks if that sequence is present
in the payload of the message. Detailed description of the Suricata
rules can be found elsewhere~\cite{suri-rule-format}. It is worth
noting that Suricata supports a total of \totoptions\ distinct options
covering \numproto\ distinct protocols. A total of 9 of these options
are for documentation.\MyComment{For instance, the purpose of rule
  option \CodeIn{msg} is to print on the output an informative message
  indicating what kind of intrusion has been observed.}
Figure~\ref{fig:distribution-rules-protocol} shows the distributions
of these options per protocol. The sum of the number of protocol
options is higher than \totoptions\ as some options are shared across
different protocols. \tname{} does not make distinction among protocols.

@@

@@

\begin{figure}[h!]
\centering
\includegraphics[scale=0.28]{figs/ICMP-packet-structure.png}
\caption{Layout of an ICMP packet.}
\label{fig:icmp-packet-layout}
\end{figure}

Rule options are often associated with specific fields of network
packets. Let's consider the ICMP protocol as an
example. Figure~\ref{fig:icmp-packet-layout} shows the layout of an
ICMP packet. The packet fields ``Flags'' and ``Time to Live'',
declared in the IP header, are mapped to the options \CodeIn{flagbits}
and \CodeIn{ttl}, respectively. Likewise, the fields ``Type'' and
``Code'', declared in the ICMP header, are mapped to the options
\CodeIn{itype} and \CodeIn{icode}. (Table~\ref{table:rules} lists some
example rule options.) The values of the options ``ttl'', ``itype'',
and ``icode'' are the same as the value of the corresponding fields in
the ICMP packet. For the field ``Flags'' the value of the
corresponding option is encoded in the bitvector. For example, if the
``Flags'' field stores the value 1010, the corresponding rule option
will be \CodeIn{flagbits:RM}, as each position of the vector encodes a
different character in the \CodeIn{flagbits} rule option. The value of
several options can be determined through transformation functions, as
those described above. One expection to this is the option
\CodeIn{threshold}, which is typically used in denial-of-service
attacks. That option operates over sequence of messages as opposed to
a single message. Consequently, it is not possible to determined its
values from a single message. More precisely, to trigger an action on
a rule containing that option a \nids\ tool needs to monitor several
messages.

@@

\Mar{----------------- estou aqui}

@@

In the following, we illustrate how \tname{} proceeds for obtaining
the rule for the SYN Flood attack. One challenge in synthesizing
rules is to discover the set of options in a rule that makes the \nids\ to
capture the malicious traffic, but allow the benign traffic to go through.  The
other parts of the rule (\ie{}, the action and the header) are either
inferred or left constant. \tname{} sets the action header to
\CodeIn{alert} and sets the header to \CodeIn{<proto> any any -> any
  any}, indicating that it infers the target protocol and uses the
most general flow description possible, \ie{}, it instructs the
\nids\ to analyze the traffic flowing from/to any address and port.
We understand that this is domain-specific information and conjecture
that the system administrator would know how to customize.

\tname{} proceeds as follows to synthesize rules for this
attack. First, it analyzes the positive traffic and extracts options
expressed in that traffic. For this example, it finds the following
set of options:

@@

\Mar{<~ Lucas/Guilheme, logo acima a gente falou que threshold nao
  poderia ser inferido a partir de uma unica mensagem e aqui a gente
  fala isto. Vcs. podem explicar o que realmente acontece?}

Note that, by construction, a rule that uses this set of options
captures the positive traffic as these options originated from the
positive traffic. Unfortunately, this rule is overspecified. It would
unlikely capture different manifestations of the same
attack. Informally, the rule includes ``too many'' characteristics of
the malicious traffic used to define it. To address this overfitting
issue, \tname{} minimizes the rule preserving the invariant that
plausible rules should only capture the positive traffic. \tname{}
potentially finds several minimal rules. The list below shows the
options for the five shortest rules it produces for this attack.

\Mar{<~ Lucas/Guilheme, vcs. podem atualizar esta informacao?}

@@

\Mar{------------------------------- estou aqui}
